LHC abort gap monitor electronics upgrade

### 2024-05-11 paper.tex

\section{Introduction}
The LHC ring stores up to \SI{350}{MJ} of beam energy, when operating at the flat top energy, spread across 2808 bunches with an average beam transverse size under \SI{1}{mm}.
%
This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}.
%
To safely extract the beam from the circulating beam lines, a series of kicker magnets belonging to the beam dump system is used. 
%
Kicker magnets need \SI{3}{\micro\second} to ramp up their magnetic field~\cite{beam_dump_system}.
%
This defines the size of an Abort Gap (AG) in the bunch train.
%
Due to system imperfections, a residual amount of particles populates this gap.
%
The typical AG population levels are two to three orders of magnitude lower in comparison to a single pilot LHC bunch.
%
When the beam dump is initiated, the particles in the AG do not receive full
extraction force, leading to losses along the dump line that could result in damage or magnet quenching.
%
%That causes the particles in the AG to
% deflect towards the vacuum chamber, possibly causing the magnets to quench. 
%
The purpose of the AGM is a~continuous monitoring of the AG population in order to trigger the abort gap cleaning when  necessary \cite{particles_in_ag} .
\\
This paper discusses the upgrade of the operational AGM installed in the LHC since 2010, detailing the system architecture and analogue electronics chain, and provides a first performance comparison  of both systems.

@@

%The particle population in the AG is measured intercepting the synchrotron radiation (SR). 
%
The AGM collects part of the extracted SR via Multi-channel plate photo-multiplier (MCP-PMT) converting it into an electric signal.
%
Neutral density filters in the optical front-end (OF) allow  changing the light intensity, thus optimizing the system dynamic range.
% 
Subsequently, the electric signal is amplified and transmitted over a \SI{100}{m}-long cable to DAQ.
%
The DAQ integrates the incoming signal in \SI{100}{\nano\second} intervals within a \SI{6.4}{\micro\second}-long acquisition window.
%
Each of the 64 \SI{100}{\nano\second} integrals is independently averaged in an Exponential Moving Averager (EMA).
%
The averaged result is published to the software with a repetition rate of \SI{10}{Hz}, which is sufficient to react to even the fastest changes in AG population.~\cite{high_sensitivity_measurement}.
%
\\
The measurement period depends on the specific MCP-PMT (HAMAMATSU R5916U-50) that features a maximum duty cycle of \SI{1}{\%}. 
%is determined by the PMT used.
%
%In our case, it is HAMAMATSU R5916U-50. 
%
%Its maximum duty cycle is \SI{1}{\%}. 
%
With a gate pulse duration of \SI{3}{\micro\second} and one LHC revolution period of \SI{89}{\micro\second}, the measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}).
%
This corresponds to a duty cycle of \SI{0.84}{\%}.
\\
%

@@

\section{System Upgrade}
To address the long-term maintenance concerns, the original system described above has to be refurbished. 
%
While the optics of the new system was copied from the operational one, the DAQ and the front-end amplifier were replaced by a modern solution.
%
A new DC-coupled amplifier was designed, thus allowing enhancing the overall dynamic range. 
%
It was decided to change from analog to digital integration of the MCP-PMT signal, via a DAQ based on  a 4-channel \SI{500}{MSPS} 14-bit FMC-form factor sampler.
%
The FMC ADC module is installed into an in-house built FPGA carrier (VFC-HD), featuring 2$\times 8$GBits DDR3 memories and a large Arria V FPGA.
%
Such a solution eliminates completely the need for the obsolete analogue integrators, as all the signal processing can be implemented in the gateware.
\\
In early 2023, the new AGM system was installed alongside the legacy system to facilitate an easy comparison of both systems' performance.
%An updated version of the AGM was in 2023 installed \textit{in parallel to the actually running AGM}, so that we can compare behavior of both systems.

@@

Figure \ref{fig:amp_schematics} depicts the block diagram of the new amplifier.
%
The amplifier was designed to match the AGM MCP-PMT and to be  highly configurable.
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))
%
A universal asynchronous receiver-transmitter (UART) interface allows for setting a trans-impedance gain.
%
The gain is adjustable in a range of \SI{5.8}{mV/\micro A} to \SI{1.1}{V/\micro A}.
%
This corresponds to a~voltage gain of \SI{41}{dB} to \SI{87}{dB} in a $50\Omega$ system.
%
A programmable DC-offset compensation is implemented in the amplifier.
%
It can be used - together with variable gain - to optimize the signal's amplitude for the ADC.
%
The amplifier design incorporates an alarm. This alarm interlocks the MCP-PMT gate when the incident light produces a current nearing the MCP-PMT's maximum allowable average current.
%The amplifier implements 
%an alarm that interlocks the MCP-PMT gate when the incident light generates a current approachng  the MCP-PMT's maximum allowable average current.
%
The alarm threshold setting is stored through the UART into the amplifier's non-volatile memory, so that 
%it protects 
the MCP-PMT is protected also after a power reset. %immediately at the amplifier power-up.
%
%The alarm status is also transmitted to the gateware.
%
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier.}
    \label{fig:amp_schematics}
\end{figure}
%
\subsection{Acquisition system}
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_daq.pdf}
    \caption{Acquisition system schematics.}
    \label{fig:bsra_daq}
\end{figure}
%
The new DAQ (Fig.~\ref{fig:bsra_daq}) receives asynchronous data samples as a stream of four frames every~\SI{8}{\nano s}, each corresponding to one channel.
%
Each frame contains 4$\times$16-bit words, storing left-aligned 14-bit ADC samples. 
%
To compute integrals over 4$\times$~\SI{25}{\nano\second} bunch slots, the system employs data tagging. 
%
The tagger resamples the LHC timing, and generates timing masks that  provide each individual ADC frame with %allowing for identification of 
bunch and turn start markers. 
%within each individual ADC frame.
%
Tags are inserted into the  2 unused least significant bits (LSBs) of each word to indicate the start of the  the bunch or turn.
\\
%
The tagged samples are stored in a~FIFO (First In, First Out), and then integrated. 
%
The integrator is triggered %begins summing values 
by the first bunch tag appearing within the acquisition window.
%
In parallel the system reads a stream of 64~values of a stored 'offset' from a pedestal memory (P[63:0]).
%
The offset is subtracted  from the integrated data stream on the fly.
%
Such pedestal subtraction is needed to remove systematic effects, e.g. related to the  MCP-PMT gate opening. 
%
The output values are filtered by EMA, the results are stored  in RAM, and then read out through the VME64x bus~\cite{my_thesis}.

>>>>

\section{Introduction}
The LHC ring stores up to \SI{350}{MJ} of beam energy, when operating at the flat top energy, spread across 2808 bunches with an average beam transverse size under \SI{1}{mm}.
%
This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}.
%
To safely extract the beam from the circulating beam lines, a~series of kicker magnets belonging to the beam dump system is used. 
%
Kicker magnets need \SI{3}{\micro\second} to ramp up their magnetic field~\cite{beam_dump_system}.
%
This defines the size of an Abort Gap (AG) in the bunch train.
%
Due to system imperfections, a residual amount of particles populates this gap.
%
The typical AG population levels are two to three orders of magnitude lower in comparison to a single pilot LHC bunch.
%
When the beam dump is initiated, the particles in the AG do not receive full
extraction force, leading to losses along the dump line that could result in damage or magnet quenching.
%
%That causes the particles in the AG to
% deflect towards the vacuum chamber, possibly causing the magnets to quench. 
%
The purpose of the AGM is a~continuous monitoring of the AG population in order to trigger the abort gap cleaning when  necessary \cite{particles_in_ag} .
\\
This paper discusses the upgrade of the operational AGM installed in the LHC since 2010, detailing the system architecture and analogue electronics chain, and provides a first performance comparison  of both systems.

@@

%The particle population in the AG is measured intercepting the synchrotron radiation (SR). 
%
The AGM collects part of the extracted SR via Multi-channel plate photo-multiplier (MCP-PMT) converting it into an electric signal.
%
Neutral density filters in the optical front-end (OF) allow  changing the light intensity, thus optimizing the system dynamic range.
% 
Subsequently, the electric signal is amplified and transmitted over a~\SI{100}{m}-long cable to DAQ.
%
The DAQ integrates the incoming signal in \SI{100}{\nano\second} intervals within a \SI{6.4}{\micro\second}-long acquisition window.
%
Each of the 64 \SI{100}{\nano\second} integrals is independently averaged in an Exponential Moving Averager (EMA).
%
The averaged result is published to the software with a repetition rate of \SI{10}{Hz}, which is sufficient to react to even the fastest changes in AG population.~\cite{high_sensitivity_measurement}.
%
\\
The measurement period depends on the specific MCP-PMT (HAMAMATSU R5916U-50) that features a maximum duty cycle of \SI{1}{\%}. 
%is determined by the PMT used.
%
%In our case, it is HAMAMATSU R5916U-50. 
%
%Its maximum duty cycle is \SI{1}{\%}. 
%
With a gate pulse duration of \SI{3}{\micro\second} and one LHC revolution period of \SI{89}{\micro\second}, the measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}).
%
This corresponds to a~duty cycle of \SI{0.84}{\%}.
\\
%

@@

\section{System Upgrade}
To address the long-term maintenance concerns, the original system described above has to be refurbished. 
%
While the optics of the new system was copied from the operational one, the DAQ and the front-end amplifier were replaced by a modern solution.
%
A new DC-coupled amplifier was designed, thus allowing enhancing the overall dynamic range. 
%
It was decided to change from analog to digital integration of the MCP-PMT signal, via a DAQ based on  a~4-channel \SI{500}{MSPS} 14-bit FMC-form factor sampler.
%
The FMC ADC module is installed into an in-house built FPGA carrier (VFC-HD), featuring 2$\times 8$GBits DDR3 memories and a large Arria V FPGA.
%
Such a solution eliminates completely the need for the obsolete analogue integrators, as all the signal processing can be implemented in the gateware.
\\
In early 2023, the new AGM system was installed alongside the legacy system to facilitate an easy comparison of both systems' performance.
%An updated version of the AGM was in 2023 installed \textit{in parallel to the actually running AGM}, so that we can compare behavior of both systems.

@@

Figure \ref{fig:amp_schematics} depicts the block diagram of the new amplifier.
%
The amplifier was designed to match the AGM MCP-PMT and to be  highly configurable.
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))
%
A universal asynchronous receiver-transmitter (UART) interface allows for setting a~trans-impedance gain.
%
The gain is adjustable in a~range of \SI{5.8}{mV/\micro A} to \SI{1.1}{V/\micro A}.
%
This corresponds to a~voltage gain of \SI{41}{dB} to \SI{87}{dB} in a $50\Omega$ system.
%
A programmable DC-offset compensation is implemented in the amplifier.
%
It can be used - together with variable gain - to optimize the signal's amplitude for the ADC.
%
The amplifier design incorporates an alarm. This alarm interlocks the MCP-PMT gate when the incident light produces a current nearing the MCP-PMT's maximum allowable average current.
%The amplifier implements 
%an alarm that interlocks the MCP-PMT gate when the incident light generates a current approachng  the MCP-PMT's maximum allowable average current.
%
The alarm threshold setting is stored through the UART into the amplifier's non-volatile memory, so that 
%it protects 
the MCP-PMT is protected also after a power reset. %immediately at the amplifier power-up.
%
%The alarm status is also transmitted to the gateware.
%
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier.}
    \label{fig:amp_schematics}
\end{figure}
%
\subsection{Acquisition system}
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_daq.pdf}
    \caption{Acquisition system schematics.}
    \label{fig:bsra_daq}
\end{figure}
%
The new DAQ (Fig.~\ref{fig:bsra_daq}) receives asynchronous data samples as a stream of four frames every~\SI{8}{\nano s}, each corresponding to one channel.
%
Each frame contains 4$\times$16-bit words, storing left-aligned 14-bit ADC samples. 
%
To compute integrals over 4$\times$~\SI{25}{\nano\second} bunch slots, the system employs data tagging. 
%
The tagger resamples the LHC timing, and generates timing masks that  provide each individual ADC frame with %allowing for identification of 
bunch and turn start markers. 
%within each individual ADC frame.
%
Tags are inserted into the  2 unused least significant bits (LSBs) of each word to indicate the start of the  the bunch or turn.
\\
%
The tagged samples are stored in a~FIFO (First In, First Out), and then integrated. 
%
The integrator is triggered %begins summing values 
by the first bunch tag appearing within the acquisition window.
%
In parallel the system reads a stream of 64~values of a stored 'offset' from a pedestal memory (P[63:0]).
%
The offset is subtracted  from the integrated data stream on the fly.
%
Such pedestal subtraction is needed to remove systematic effects, e.g. related to the  MCP-PMT gate opening. 
%
The output values are filtered by EMA, the results are stored  in RAM, and then read out through the VME64x bus~\cite{my_thesis}.
